An Exploration of Severe Weather Events: NOAA Storm Data, 1950 - 2011
=====================================================================

## Synopsis

summary analysis / results; no more than 10 sentences

## Data Processing

### Preprocessing

Before processing and analysis, we must first obtain the data:

```{r, results='hide'}
## set WD on local machine to the location of this Rmd file; setup data dir
setwd("C:/Users/570815/Dropbox/Coursera/R Working Directory/RepData_PeerAssessment2")
zipURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
if(!file.exists("./data")) {dir.create("./data")}
```

To download on Windows [try() wrapper is to suppress superfluous Error message
produced when knitting]:

```{r, results='hide', warning=FALSE}
try(download.file(zipURL, destfile="./data/NOAAdata.csv.bz2"))
dateDownloaded <- date()
```

To download on Mac:

```{r, results='hide', warning=FALSE}
download.file(zipURL, destfile="./data/NOAAdata.csv.bz2", method="curl")
dateDownloaded <- date()
```

After downloading is complete, we can read the data into R. Testing the connection 
to the bz2 file, we see that the values are conveniently comma-separated:

```{r}
testcon <- bzfile("./data/NOAAdata.csv.bz2")
open(testcon)
readLines(testcon, n=3)
close(testcon)
```

As this is the case, we can use read.csv() to read the data into R:

```{r, cache=TRUE}
## takes a while to read - roughly 33 million data points
sd <- read.csv("./data/NOAAdata.csv.bz2")
```

The storm data are now loaded and can be processed for analysis.

### Processing for Analysis

There are two key questions we are trying to answer in our analysis:

1. Across the United States, which types of events (as indicated in the EVTYPE 
variable) are most harmful with respect to population health?

2. Across the United States, which types of events have the greatest economic 
consequences?

We will begin our analysis with the first question.  

## Results

figures can go here; must have at least 1 figure containing a plot, and no more than 3


## Wrap-up / final notes

maybe?





