An Exploration of Severe Weather Events: NOAA Data, 1950 - 2011
===============================================================

## Synopsis

summary analysis / results; no more than 10 sentences

## Data Processing

Before processing and analysis, we must first obtain the data:

```{r, results='hide'}
## set WD on local machine to the location of this Rmd file; setup data dir
setwd("C:/Users/570815/Dropbox/Coursera/R Working Directory/RepData_PeerAssessment2")
zipURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
if(!file.exists("./data")) {dir.create("./data")}
```

To download on Windows [try() wrapper is to suppress superfluous Error message
produced when knitting]:

```{r, results='hide', warning=FALSE}
try(download.file(zipURL, destfile="./data/NOAAdata.zip"))
dateDownloaded <- date()
```

To download on Mac:

```{r, results='hide', warning=FALSE}
download.file(zipURL, destfile="./data/AMdata.zip", method="curl")
dateDownloaded <- date()
```

After downloading is complete, extract the NOAA files to the ./data
directory in the working directory, and load the data into R:

```{r}

```

The data are now loaded and can be processed.

### Processing: doin' thangs

stuff goes here for processing


## Results

figures can go here; must have at least 1 figure containing a plot, and no more than 3


## Wrap-up / final notes

maybe?





