An Exploration of Severe Weather Events: NOAA Storm Data, 1950 - 2011
=====================================================================

## Synopsis

summary analysis / results; no more than 10 sentences

## Data Processing

### Preprocessing

Before processing and analysis, we must first obtain the data:

```{r, results='hide'}
## set WD on local machine to the location of this Rmd file; setup data dir
setwd("C:/Users/570815/Dropbox/Coursera/R Working Directory/RepData_PeerAssessment2")
zipURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
if(!file.exists("./data")) {dir.create("./data")}
```

To download on Windows [try() wrapper is to suppress superfluous Error message
produced when knitting]:

```{r, results='hide', warning=FALSE}
try(download.file(zipURL, destfile="./data/NOAAdata.csv.bz2"))
dateDownloaded <- date()
```

To download on Mac:

```{r, results='hide', warning=FALSE}
download.file(zipURL, destfile="./data/NOAAdata.csv.bz2", method="curl")
dateDownloaded <- date()
```

After downloading is complete, we can read the data into R. Testing the connection 
to the bz2 file, we see that the values are conveniently comma-separated:

```{r}
testcon <- bzfile("./data/NOAAdata.csv.bz2")
open(testcon)
readLines(testcon, n=3)
close(testcon)
```

As this is the case, we can use read.csv() to read the data into R:

```{r, cache=TRUE}
## takes a while to read - roughly 33 million data points
sd <- read.csv("./data/NOAAdata.csv.bz2")
```

The storm data are now loaded and can be processed for analysis.

### Processing for Analysis

There are two key questions we are trying to answer in our analysis:

1. Across the United States, which types of events (as indicated in the EVTYPE 
variable) are most harmful with respect to population health?

2. Across the United States, which types of events have the greatest economic 
consequences?

We will begin our analysis with the first question. The variables of interest to 
us in the dataset for looking at population health are fatalities and injuries. 
The following code obtains the total number of injuries and fatalities by event 
type:

```{r, cache=TRUE}
library(plyr)
## may take a little bit of time to execute the ddply operation
healthDF <- ddply(sd, .(EVTYPE), summarize, injuries=sum(INJURIES, na.rm=T), 
                  fatalities=sum(FATALITIES, na.rm=T))
```

We can quickly examine some of the "most harmful" events as categorized by 
causing injury and/or fatality:

```{r}
healthDF <- as.data.frame(healthDF)
## top 5 events causing injuries:
topInjury <- healthDF[with(healthDF, order(injuries, fatalities, decreasing=TRUE)),]
head(topInjury, 5)
## top 5 events causing fatalities:
topFatality <- healthDF[with(healthDF, order(fatalities, injuries, decreasing=TRUE)),]
head(topFatality, 5)
```

Can create a plot of the top N things - maybe like a 2 color histogram...

get merged list of top 20 things (arbitrary choice) -- get, say, top 50 EVTYPEs for each of injuries and fatalities, merge the uniques (in order), then truncate to 20

then get a subset of the healthDF data where EVTYPE %in% merged list

plot the total recorded injuries and fatalities for each event type on a bar chart -- need to figure out how to do a double bar chart.  i think that this would be a decent, if rough, "analysis" of the "most harmful things"

Now we turn to the second question.

(similar analysis but doing a subset of the data related to economic damage variables -- looks like PROPDMG and CROPDMG - the niggle here is that the <col>EXP column gives you the units of the <col>, so if PROPDMG = 25 and PROPDMGEXP = K, the economic damage is $25,000)

```{r}
#PROBABLY WANT TO CACHE THIS CHUNK

# maybe first do some looping action somehow to get combined damage $ figures ?
#econDF <- a subset of sd on select variables, similar to healthDF
```

once i hvae these data figured out, can do some analysis and come up with a useful sort of plot; maybe just as simple as a version of the other one OR maybe better to do a panel plot to mix things up / do a slightly different type of analysis


## Results

figures can go here; must have at least 1 figure containing a plot, and no more than 3

(maybe keep processing code up above to get healthDF / econDF datasets, then do plotting code here + written explanation of what we're looking at in the plots)

## Wrap-up / final notes

maybe?





