An Exploration of Severe Weather Events: NOAA Storm Data, 1950 - 2011
=====================================================================

## Synopsis

summary analysis / results; no more than 10 sentences

## Data Processing

### Preprocessing

Before processing and analysis, we must first obtain the data:

```{r, results='hide'}
## set WD on local machine to the location of this Rmd file; setup data dir
setwd("C:/Users/570815/Dropbox/Coursera/R Working Directory/RepData_PeerAssessment2")
zipURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
if(!file.exists("./data")) {dir.create("./data")}
```

To download on Windows [try() wrapper is to suppress superfluous Error message
produced when knitting]:

```{r, results='hide', warning=FALSE}
try(download.file(zipURL, destfile="./data/NOAAdata.csv.bz2"))
dateDownloaded <- date()
```

To download on Mac:

```{r, results='hide', warning=FALSE}
download.file(zipURL, destfile="./data/NOAAdata.csv.bz2", method="curl")
dateDownloaded <- date()
```

After downloading is complete, we can read the data into R. Testing the connection 
to the bz2 file, we see that the values are conveniently comma-separated:

```{r}
testcon <- bzfile("./data/NOAAdata.csv.bz2")
open(testcon)
readLines(testcon, n=3)
close(testcon)
```

As this is the case, we can use read.csv() to read the data into R:

```{r, cache=TRUE}
## takes a while to read - roughly 33 million data points
sd <- read.csv("./data/NOAAdata.csv.bz2")
```

The storm data are now loaded and can be processed for analysis.

### Processing for Analysis

There are two key questions we are trying to answer in our analysis:

1. Across the United States, which types of events (as indicated in the EVTYPE 
variable) are most harmful with respect to population health?

2. Across the United States, which types of events have the greatest economic 
consequences?

We will begin our analysis with the first question. The variables of interest to 
us in the dataset for looking at population health are fatalities and injuries. 
The following code obtains the total number of injuries and fatalities by event 
type:

```{r, cache=TRUE}
library(plyr)
## may take a little bit of time to execute the ddply operation
healthDF <- ddply(sd, .(EVTYPE), summarize, injuries=sum(INJURIES, na.rm=T), 
                  fatalities=sum(FATALITIES, na.rm=T))
```

We can quickly examine some of the "most harmful" events as categorized by 
causing injury and/or fatality, noting that a few of the top items are the same:

```{r}
healthDF <- as.data.frame(healthDF)
## top 5 events causing injuries:
topInjury <- healthDF[with(healthDF, order(injuries, fatalities, decreasing=TRUE)),]
head(topInjury, 5)
## top 5 events causing fatalities:
topFatality <- healthDF[with(healthDF, order(fatalities, injuries, decreasing=TRUE)),]
head(topFatality, 5)
```

To get a better sense of the most harmful events to human health, we can do a 
bit more processing of the injury and fatality data to obtain a shared set of 
events, which we will plot in the Results section below:

```{r}
## obtain the top 25 most-harmful effects as indicated by injuries / fatalities
top25i <- as.character(topInjury$EVTYPE[1:25])
top25f <- as.character(topFatality$EVTYPE[1:25])
## take the unique values from these top lists
topEventNames <- unique(c(top25i, top25f))
## subset the health effects events on this set of most-harmful events
topHealthEvents <- healthDF[healthDF$EVTYPE %in% as.factor(topEventNames),]
```

Now we turn to the second question.

(similar analysis but doing a subset of the data related to economic damage variables -- looks like PROPDMG and CROPDMG - the niggle here is that the <col>EXP column gives you the units of the <col>, so if PROPDMG = 25 and PROPDMGEXP = K, the economic damage is $25,000)

```{r}
#PROBABLY WANT TO CACHE THIS CHUNK

# maybe first do some looping action somehow to get combined damage $ figures ?
#econDF <- a subset of sd on select variables, similar to healthDF
```

once i hvae these data figured out, can do some analysis and come up with a useful sort of plot; maybe just as simple as a version of the other one OR maybe better to do a panel plot to mix things up / do a slightly different type of analysis


## Results

### Health-affecting Events

Based on the analysis of health effects caused by weather events in the prior section, 
we can construct the following plot of the weather events that are most harmful to human 
health, to give a clearer sense of which events we might want to focus our efforts on 
in terms of emergency preparation. Note that the y-axis on this graph is plotted on a 
logarithmic scale, as Tornado injuries were significantly greater in magnitude than all 
other event injuries.

```{r, message=FALSE, fig.height=11, fig.width=11}
library(reshape2)
library(ggplot2)
moltenHealth <- melt(topHealthEvents, id.vars=("EVTYPE"))
ggplot(moltenHealth, aes(x=EVTYPE, y=value)) + geom_bar(aes(fill=variable), position="dodge") + scale_y_log10() + theme(axis.text.x = element_text(size=10, angle=90, hjust=1, vjust=0.25)) + labs(title="Top Health-affecting Weather Events", x="Event Type", y="Log number of injuries and fatalities") + scale_fill_discrete(name="Health Effect")
```

Looking at this plot, we can make a few observations. While the most harmful weather 
events typically cause more injuries than fatalities, those related to cold and wind chill 
actually cause more fatalities than injuries -- priority remediation activities may need 
to be devised for these events that are generally more fatal than simply injurious. 

The most harmful event types overall can also be clearly seen here -- tornadoes, heat, 
floods, lightning, and winter storms (this lattermost likely to be associated with the 
cold and wind chill events as well) are among the events with the greatest negative 
impact on human health.

### Economy-affecting Events


figures can go here; must have at least 1 figure containing a plot, and no more than 3

(maybe keep processing code up above to get healthDF / econDF datasets, then do plotting code here + written explanation of what we're looking at in the plots)

## Wrap-up / final notes

maybe?





